---
title: "Module 3"
author: "UNK Biology"
---

# Learning objectives:

To continue gaining familiarity with the R statistical framework by using RStudio to:

-   Calculate mean and standard deviation

-   Calculate median, 25th and 75th percentiles, and interquartile range

-   Use functions in `ggplot2` (part of `tidyverse`) to create boxplots, bar charts, scatterplots, and line graphs.

# Packages

In this problem set, we will use more functions associated with the `tidyverse` package, including a subset of this package - `ggplot2` - which is used for creating publication-quality graphs.

First, we need to load this package. Remember - we installed this in a previous module, so we should not need to install in again and should be able to simply load it.

```{r}
library(tidyverse)
```

# Datasets:

We will be working with two real-world datasets this week that are pulled directly from Zenodo, the European Union's open-source repository for scientific data. We will be using the `read_csv` command from `tidyverse` to read files directly from the internet into *R*.

## Dataset 1

Our first dataset looks at the energetic costs associated with nesting in an Arctic seabird, the [Common Eider *Somateria mollissima*](https://ebird.org/species/comeid) [@hoyvik_hilde_data_2017]. These data represent the open-access permanent archive of a publication based on these data [@hoyvik_hilde_mind_2016]. The authors constructed artificial shelters around some nests to understand how it affected the physical condition of incubating female eiders. *Note* that I do a few steps below to (1) reduce the size of the data file and (2) rename columns to be more intuitive. All of these steps involve the `tidyverse` steps we've discussed before.

```{r}
bird_data <- read_csv("https://zenodo.org/records/4960831/files/Data%20file.csv") %>% 
  # select columns of interest
  select(`Nest ID`, Ring, MeanT, MeanW, MeanH, `Differ. Days`, 
         Mass, `Mass recapture`, CS, CSRecapture, `Shelter added (Y/N)`) %>% 
  # rename columns
  rename(Nest_ID = `Nest ID`, Bird_ID = Ring,
         Mean_Temp = MeanT, Mean_Wind = MeanW, 
         Mean_Humid = MeanH, Days_between_captures = `Differ. Days`,
         Mass_Initial = Mass, Mass_Recapture = `Mass recapture`,
         Clutch_size_initial = CS, Clutch_size_recapture = CSRecapture,
         Shelter_added = `Shelter added (Y/N)`)
```

## Dataset 2

Our second dataset relates to the effects of nitrogen on the growth of Scots Pine and Norway Spruce [@jetsonen_volume_2025]. This is a dataset related to a publication on these trees and the volume they add that came out a year prior, and represents another open-access dataset [@jetsonen_effects_2024]. Like [dataset 1], we are going to select specific columns and rename them for this assignment.

```{r}
tree_data <- read_csv("https://zenodo.org/records/14733005/files/N_fertilization_Finland_data.csv") %>% 
  select(tree, v_incr_annual_compare, FN, Pa, Tsum, SF) %>% 
  rename(Tree_Species = tree, 
         Annual_Mean_Volume_Growth = v_incr_annual_compare,
         Nitrogen_Dose = FN, Mean_Annual_Precipitation = Pa,
         Long_term_temperature_sum = Tsum, time_since_fertilization = SF)
```

# Procedure:

1.  Watch Module 3: Describing and Visualizing Data in R for a walk-through of these initial procedures.
2.  Open a new *R* script or *Rmarkdown* script in the source pane by clicking on the `File menu > New File > R Script` or `Rmarkdown`.
3.  If necessary, you can clear the environment and plots panes from *R*’s memory by type `rm(list=ls())` in the code pane and then hitting `CTRL` + `L` to clear the coding pane's text. Alternatively, this can be accomplished by clicking the broom icon at the top of both of these panes and clear the console by clicking on `Editmenu > Clear Console`.
4.  Set your working directory and import both datasets. Review Modules 1 and 2 procedures, if necessary, for a reminder of how to complete these tasks.

-   For these procedures and video, [dataset 1] was imported as a dataframe object named `bird_data`, and [dataset 2] as a dataframe object named `tree_data`.

Descriptive statistics refers to simple calculations that summarize or describe a given variable. Measures of central tendency commonly used are the mean (i.e., average) and median (i.e, 50th percentile). Corresponding measures of spread are standard deviation (accompanying the mean) and interquartile range or 25th and 75th percentiles (accompanying the median).

## Section A. Calculate mean and standard deviation

5.  Create a new column for the difference in mass as a percent of the birds' initial mass using the command `mutate`.

```{r}
bird_data <- bird_data %>% 
  mutate(Percent_mass_change = (Mass_Recapture - Mass_Initial)/Mass_Initial)
```

6.  Calculate the mean and standard deviation of the `Percent_mass_change` variable separately for whether the nest was or was not sheltered (`Shelter_added`; `Y` or `N`). First, we will calculate these "by hand" using the following equations:

$$
\bar{x}=\frac{\Sigma_1^n x}{n}
$$

$$
s^2 = \frac{\Sigma_i^n (x_i - \bar{x})^2}{n}
$$

In order to do this, we will need to write a `function`. A `function` is a specific *R* command that creates a pipeline for some sort of input data.

In writing these functions, it's important to know that we can perform subtraction, multiplication, and other mathematical functions across multiple values as once. For example, if we want to know what $1$ through $10$ all $-5.5$ are equal, we can do the following:

```{r}
# denote multiple numbers with :
x <- 1:10

x - 5.5
```

We can write our own functions in *R* that will perform these functions as follows:

```{r}
our_mean <- function(x){
  # get sum of input
  sum_x <- sum(x)
  # calculate length of input
  n <- length(x)
  # calculate mean
  xbar <- sum_x/n
  # return mean to console
  # necessary for saving outside function
  return(xbar)
}

our_sd <- function(x){
  # need to calculate mean first!
  # copied from above
  # get sum of input
  sum_x <- sum(x)
  # calculate length of input
  n <- length(x)
  # calculate mean
  xbar <- sum_x/n
  
  # calculate deviants
  deviants <- x - xbar
  deviants_sq <- deviants^2
  our_variance <- sum(deviants_sq)/n
  our_sd <- sqrt(our_variance)
  # return answer
  return(our_sd)
}
```

7.  Remember, we can use `select` to get a specific set of data from a `dataframe` in *R*. I will calculate these metrics for one set of variables here:

```{r}
# filter data
no_shelter_change <- bird_data %>% 
  # get non-sheltered birds
  filter(Shelter_added == "N")

# extract data of interest ONLY
no_shelter_mass_change <- no_shelter_change$Percent_mass_change

# view first few values
head(no_shelter_mass_change)
```

8.  Now that we have our data subsetted, we can calculate "by hand".

```{r}
our_mean(no_shelter_mass_change)
```

```{r}
our_sd(no_shelter_mass_change)
```

8.  Now, calculate the following for individuals *with* shelters. Your answers should be as follows (I have hidden my subsetting code):

```{r, echo = F}
# filter data
shelter_change <- bird_data %>% 
  # get non-sheltered birds
  filter(Shelter_added == "Y")

# extract data of interest ONLY
shelter_mass_change <- shelter_change$Percent_mass_change
```

```{r}
our_mean(shelter_mass_change)
```

```{r}
our_sd(shelter_mass_change)
```

9.  Now, let's compare our answers to those calculated by *R*. These values may differ slightly based on rounding in *R*, but should be very close. Remember - you can use the same objects that we used before if they are still stored in your memory!

```{r}
mean(no_shelter_mass_change)
```

```{r}
sd(no_shelter_mass_change)
```

10. Now, we want to run `pastecs` command `stat.desc` on part of the dataframe (for example, the mass change with a shelter).

```{r}
# use library to load package
# this may not be shown in future modules!
library(pastecs)

# stat.desc, rounded to 2 decimals
stat.desc(shelter_mass_change) %>% 
  round(2)
```

Run the command on the mass data for birds with no shelter as well.

11. Compare the weight data for birds with a shelter and without a shelter. What similarities do you see? What differences do you see?

## Section B. Calculate the median (50th), 25th, and 75th percentiles and interquartile range (IQR) {#section-b.-calculate-the-median-50th-25th-and-75th-percentiles-and-interquartile-range-iqr}

Quantiles are a way of looking at the response value at which a given percentage of observations are captured. The most common quantiles are the 25th, 50th, and 75th percentiles. This specific set of quantiles are also called quartiles: 1st quartile (25th percentile), median (50th percentile), and 3rd quartile (75th percentile). The interquartile range (IQR) measures spread representing the middle 50% of observations; that is, 25% of observations fall below the 25th percentile and 25% of observations are above the 75th percentile.

The following examples will use the data for mass change with no shelter, which I called `no_shelter_mass_change` above.

11. First, we will be calculating the median. We can do this with the command `median`.

```{r}
median(no_shelter_mass_change)
```

12. Next, we will calculate the quantiles. We will do this using `quantile`, but note that this function includes multiple methods.

```{r}
quantile(no_shelter_mass_change)
```

If we want a specific quantile, we can state it.

```{r}
quantile(no_shelter_mass_change,
         probs = 0.25)
```

There are nine different algorithms for calculating quantiles. We will be using `type 6` in this class, which must be specified.

```{r}
quantile(no_shelter_mass_change,
         probs = 0.25,
         type = 6)
```

***Note*** that the above 25th percentile is different that the default method! Specifying a type is important, and **in this class use type 6**. Type 6 is the same method as is used by *Minitab* and *SPSS*, other statistical softwares, and calculates the quantiles using the equation $p_k = \frac{k}{n+1}$.

Just like other functions, we can save a quantile as an object as well.

```{r}
q25 <- quantile(no_shelter_mass_change,
                probs = 0.25, 
                type = 6)

q25 %>% round(2)
```

13. Calculate the interquartile range. We can do this using the command `IQR`. The interquartile range is the distance between the 25th and 75th percentiles.

```{r}
IQR(no_shelter_mass_change)
```

***NOTE*** that just like `quantile`, `IQR` has different methods and we should use type 6.

```{r}
IQR(no_shelter_mass_change,
    type = 6)
```

## Section C. Use `ggplot` to create plots

There are multiple ways to graph data in *R*, to illustrate we can compare histograms produced using the `hist()` function and using function `ggplot()` in the `ggplot2` package. ***NOTE*** that if you have `tidyverse` installed, as you should already, that you already have `ggplot2` installed! The `tidyverse` is actually a set of multiple packages that all get installed together, of which `ggplot2` is one major component.

Here, we are going to be focusing on `geom_histogram` which creates histograms, but there are also ways to produce many other plots, including `geom_boxplot`, `geom_bar`, `geom_point`, and `geom_line`.

14. For each mass group, with and without shelter, we will create a histogram using `hist`.

```{r}
hist(shelter_mass_change)
```

```{r}
hist(no_shelter_mass_change)
```

15. Now, we want to create a histogram in `ggplot2`. This is a powerful plotting function, which can be confusing but will create publication quality plots with the right commands. Several parts of the general function syntax are outlined below.

-   `ggplot`: the base function, in which you list the dataframe to be plotted. **The input must be an object with class `dataframe`**.

    -   `aes`: subargument, where aesthetics are listed.

-   `geom_histogram`: create a histogram

-   `xlab`: x axis label

-   `ylab`: y axis label

-   `ggtitle`: main plot title

-   Themes can be used to change plot appearance and make it more readable.

    -   `theme_classic`: nice plain plot with no background.

    -   `theme_minimal`: clean background but with major grid lines shown.

First, we will create a basic histogram, and then we will make a "fancy" histogram.

```{r}
ggplot(bird_data,
       aes(x = Percent_mass_change)) +
  geom_histogram()
```

Above, we can see that this histogram is pretty similar to a regular histogram from the `hist` command. Now, let's make it look nicer, and even parse out birds based on other variables. The point of each command is noted below. You may see an error about bin widths; this can be ignored.

```{r}
# define dataset to plot
ggplot(bird_data,
       # described aesthetics; thing to plot
       aes(x = Percent_mass_change,
           # thing to group by
           # for stratified studies
           fill = Shelter_added)) +
  # create a histogram
  # alpha sets transparency level
  geom_histogram(alpha = 0.6, 
                 # identity prevents "stacking" of bars
                 position = "identity") +
  # change x label
  xlab("Percent mass change") +
  # change y label
  ylab("Number of individuals") +
  # change legend label
  labs(fill = "Shelter added") +
  # set minimal theme
  theme_minimal()
```

We can also do a density histogram by changing `geom_histogram` to `geom_density`. A density histogram is *smoothed* and shows the *percent* of individuals in each category.

```{r}
ggplot(bird_data,
       aes(x = Percent_mass_change,
           fill = Shelter_added)) +
  ### Change to density ###
  geom_density(alpha = 0.6, 
                 position = "identity") +
  ### Change to percent from count ###
  xlab("Percent mass change") +
  ylab("Percent of individuals") +
  labs(fill = "Shelter added") +
  theme_minimal()
```

16. Above, we saw an error related to "bin count". We can set bin count manually by using the argument `bins` in `geom_histogram`. Below is an example that is the same as shown previously but with `bins = 6`.

```{r}
# define dataset to plot
ggplot(bird_data,
       # described aesthetics; thing to plot
       aes(x = Percent_mass_change,
           # thing to group by
           # for stratified studies
           fill = Shelter_added)) +
  # create a histogram
  # alpha sets transparency level
  geom_histogram(alpha = 0.6, 
                 # identity prevents "stacking" of bars
                 position = "identity",
                 # set bin number
                 bins = 6) +
  # change x label
  xlab("Percent mass change") +
  # change y label
  ylab("Number of individuals") +
  # change legend label
  labs(fill = "Shelter added") +
  # set minimal theme
  theme_minimal()
```

## Section E. Create a boxplot using `ggplot2`

17. Create a boxplot for the column `Percent_mass_change` from the `bird_data` dataset. The syntax for this is:

```{r, eval = F}
### THIS CODE WILL NOT RUN ###
# Example of set up
ggplot(dataframe, 
       # x is grouping variable; like condition
       # X is not needed, can be blank
       aes(x = x.variable,
           # variable to "boxplot"
           y = y.variable)) +
  geom_boxplot()
```

If you enter your correctly, you should get an output like this:

```{r, echo = F}
ggplot(bird_data, 
       aes(y = Percent_mass_change)) +
  geom_boxplot()
```

If you add a grouping variable, like `x = Shelter_added` to the `aes` argument, you will get something like this:

```{r}
ggplot(bird_data, 
       aes(x = Shelter_added, 
           y = Percent_mass_change)) +
  geom_boxplot()
```

18. Like we did before, try changing the names of the axes, and try adding the command `fill = Shelter_added` to see what happens. Remember, you can also add things like `theme_minimal()` to make the plot "tidier". See previous plots for formatting. Here is an example with axes renamed and fill added.

```{r,echo = F}
ggplot(bird_data, 
       aes(x = Shelter_added, 
           y = Percent_mass_change,
           fill = Shelter_added)) +
  geom_boxplot() +
  xlab("Shelter Added") +
  ylab("Percent Mass Change") +
  theme_minimal()
```

## Section F. Create a bar chart using `ggplot2`

19. Next, we are going to create a bar chart for the `Percent_mass_change` data. This is fundamentally similar to the above plots, but instead of `geom_boxplot` or any other argument, we will do the following:

-   `geom_bar(stat = "summary", fun = "mean")`

You should get a plot that looks like the following:

```{r}
ggplot(bird_data, 
       aes(x = Shelter_added,
           y = Percent_mass_change,
           fill = Shelter_added)) +
  geom_bar(stat = "summary",
               fun = "mean")
```

## Section G. Create an XY scatterplot using `ggplot2`

20. For this scatterplot, we are going to compare `Mass_Initial` to `Mass_Recapture`. We will use `Mass_Initial` as our independent variable (`x`), and `Mass_Recapture` as our dependent variable (`y`).

-   For this plot, we are using `geom_point`, which calls a scatterplot of points.

-   For this one, we are also going to show how sections of this plot can be saved as objects.

```{r}
a1 <- ggplot(bird_data, 
             aes(x = Mass_Initial,
                 y = Mass_Recapture))
a2 <- geom_point()
a3 <- theme_minimal()
```

Try these two combinations to see how you can add things in piecemeal:

```{r}
a1 + a2
```

```{r}
a1 + a2 + a3
```

We can also split this up by color as well. All we have to do is change the `a1` object above to include `colour`, which is similar to `fill` but better suited for this plot.

```{r}
a1 <- ggplot(bird_data,
             aes(x = Mass_Initial,
                 y = Mass_Recapture,
                 colour = Shelter_added))

a1 + a2 + a3
```

## Section H. Create a line graph using `ggplot2`

21. Next, we will create a line graph. This is essentially the same as the previous steps, except we will be using the argument `geom_line` to create our plot.

```{r}
ggplot(bird_data, aes(x = Mass_Initial,
                      y = Mass_Recapture,
                      colour = Shelter_added)) +
  geom_line()
```

***Notice*** that this is a fairly messy plot, but it still works and does a good job of illustrating how these things relate. We can still change the axes labels with commands like `xlab` and `ylab`.

# Problem Set 3

The following questions pertain to the `tree_data` object.

## (4 points) Using the `tree_data` object, split the dataset up by `Tree_Species` and Calculate the mean, standard deviation, median, and interquartile range for the column `Annual_Mean_Volume_Growth`. Refer to [Section A. Calculate mean and standard deviation] and [Section B. Calculate the median (50th), 25th, and 75th percentiles and interquartile range (IQR)](#section-b.-calculate-the-median-50th-25th-and-75th-percentiles-and-interquartile-range-iqr) for more information on performing these analyses.

***HINT***: There are only two "species" of tree - `pine` and `spruce`.

**REMEMBER** - answers should be rounded to the nearest tenth (0.1) as this is the precision of the available data.

| Measurement         | `pine` | `spruce` |
|---------------------|--------|----------|
| Mean                |        |          |
| Standard Deviation  |        |          |
| Median              |        |          |
| Interquartile Range |        |          |

: Example table for organizing results.

## (2 points) Use package `ggplot2` to create one boxplot with `Annual_Mean_Volume_Growth` on the *y*-axis and each `Tree_Species` on the *x*-axis. Be sure axes are clearly labeled and include units of measure if applicable. You will be asked to insert or copy-paste image files or screenshots of your histograms in *Canvas*.

## (2 points) Use package `ggplot2` to create one bar chart with `Annual_Mean_Volume_Growth` on the *y*-axis and each `Tree_Species` on the *x*-axis. Be sure axes are clearly labeled and include units of measure if applicable. You will be asked to insert or copy-paste image files or screenshots of your histograms in *Canvas*.

## (2 points) Use the `tree_data` dataset and package `ggplot2` to create a scatterplot with `Nitrogen_Dose` on the *x*-axis and `Annual_Mean_Volume_Growth` max on the *y*-axis. Be sure axes are clearly labeled and include units of measure if applicable. You will be asked to insert or copy- paste image files or screenshots of your histograms in *Canvas*.

## (2 points) Use package `ggplot2` to create a line graph with `Nitrogen_Dose` max on the *x*-axis and `Annual_Mean_Volume_Growth` on the y-axis. There should be one line for each `Tree_Species` and lines should be different colors. Be sure axes are clearly labeled and include units of measure if applicable. You will be asked to insert or copy-paste image files or screenshots of your histograms in Canvas.

## (BONUS +1 point possible) Putting together what you learned, create a scatterplot from the `tree_data` dataset in which `Nitrogen_Dose` is on the x-axis, `Annual_Mean_Volume_growth` is on the y-axis, and points are grouped and colored by `Tree_Species`. What do you notice about the graph?

## (BONUS +1 point possible) In the code used for the previous question, add the command `geom_smooth(method = 'lm')`. Rerun the ggplot() function. What do you notice about how the graph changed?

## (BONUS +1 point possible) Recreate the bar chart from the previous problem set question. Run the code again, but first add `position = "dodge"` to the `geom_bar()` function. Be sure to use a comma to separate it from other attributes. Be sure axes are clearly labeled and include units of measure if applicable. You will be asked to insert or copy-paste image files or screenshots in *Canvas*. How did the graph change this time?

## For the possibility of partial credit, compile your work in a single document and upload it in Canvas.

# Submitting your work for the Module 3 Problem Set Assignment:

You do not need to answer any questions included in the procedures sections, those are for your self-reflection only. Questions in *Canvas* may be worded slightly differently than here, usually for brevity, but with the same meaning. If there is a discrepancy in what is being asked between this document and the question form in *Canvas*, answer based on what is asked in this document. Also, please let me know as soon as possible so I can get it fixed. Submit your work on or before the due date indicated on the assignment page.
